Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
some work on related work and contributions

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of long-existing CPU features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in one implementation
  of Python and is then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

Eliminate GIL:
\begin{itemize}
\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
\item Previous attempts with STM: \cite{stmupdate13}
\end{itemize}

Similar STMs:
\begin{itemize}
\item FastLane: \cite{warmhoff13}
\item TML: \cite{spear09}
\item Virtualizing HTM: \cite{rajwar05}
\item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
  detection, otherwise hardware extensions required; assumes most
  transactions fit HTM capacities (not so true here); COW using page-faults;
  they assume OS-level access to page-tables (maybe not inherent to their
  design); eval on simulator; value-based confl detection;

 (XTM can be
  implemented either in the OS as part of the virtual memory manager or
  between underlying TM systems and the OS, like virtual machines;
  Conflicts for overflowed transactions are tracked at page granularity;
  XTM-e allows conflict detection at cache line granu-
  larity, even for overflowed data in virtual memory)
\item using mmap(): Memory-Mapped Transactions
\item mem-protected conflict detection: \cite{martin09}
\end{itemize}

>>>>

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well on low numbers of CPUs.
\item We make novel use of the memory management capabilities of
  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
\item We integrate our STM system closely with a garbage collector
  (GC) in order to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

There have been several attempts at removing the GIL using TM. We
ourselves did a few earlier attempts with more conventional STM
systems~\cite{stmupdate13}. Because of the huge overhead we
encountered, the result was mostly useless on $\le 4$ CPU cores.
Since this could be considered the main operating conditions of
dynamic languages, it was never as useful as we wanted it to be.

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent \cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary atomic blocks, let
us to believe that STM is the right choice at the moment.

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of \cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified in our case.

From the other end, there also have been attempts at lowering the
overhead of STM~\cite{warmhoff13,spear09} especially for a low number
of CPU cores. They have completely different approaches and it would
be interesting to see how well they work for our specific use
case. \remi{enough?}

Similarly to us, \cite{martin09} use memory management techniques to
provide isolation between transactions. They focus on providing strong
atomicity between transactional and non-transactional code.  We,
however, use inevitable transactions to provide strong atomicity.
Additionally, while they use page fault handlers to detect conflicts,
we use traditional write barriers. Again, there are only some
superficial similarities between our approach and theirs.

% Similar STMs:
% \begin{itemize}
% \item FastLane: \cite{warmhoff13}
% \item TML: \cite{spear09}
% \item Virtualizing HTM: \cite{rajwar05}
% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
%   detection, otherwise hardware extensions required; assumes most
%   transactions fit HTM capacities (not so true here); COW using page-faults;
%   they assume OS-level access to page-tables (maybe not inherent to their
%   design); eval on simulator; value-based confl detection;

%  (XTM can be
%   implemented either in the OS as part of the virtual memory manager or
%   between underlying TM systems and the OS, like virtual machines;
%   Conflicts for overflowed transactions are tracked at page granularity;
%   XTM-e allows conflict detection at cache line granu-
%   larity, even for overflowed data in virtual memory)
% \item using mmap(): Memory-Mapped Transactions
% \item mem-protected conflict detection: \cite{martin09}
% \end{itemize}
